Information-theoretic measures for non-linear causality detection: \\ application to social media sentiment and cryptocurrency prices

### 2020-07-24 paper/submission.tex
Attempt to clarify terminology re: causality vs statistical/Granger causality

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
    Information transfer between time series is calculated using the asymmetric information-theoretic measure known as transfer entropy. Geweke's autoregressive formulation of Granger causality is used to compute linear transfer entropy, and Schreiber's general, non-parametric, information-theoretic formulation is used to quantify non-linear transfer entropy. 
    We first validate these measures against synthetic data. Then we apply these measures to detect causality between social sentiment and cryptocurrency prices. We validate results by performing permutation tests by shuffling the time series, and calculate the Z-score. We also investigate different approaches for partitioning in nonparametric density estimation which can improve the significance. 
    Using these techniques on sentiment and price data over a 48-month period to August 2018, for four major cryptocurrencies, namely bitcoin (BTC), ripple (XRP), litecoin (LTC) and ethereum (ETH), we detect significant information transfer, on hourly timescales, with greater net information transfer from sentiment to price for XRP and LTC, and instead from price to sentiment for BTC and ETH. We report the scale of non-linear causality to be an order of magnitude larger than linear causality.
\end{abstract}

@@

  Though developed relatively recently, information-theoretic methods have been used with success in research across disciplines, to detect information transfer where interventionist approaches are not possible. For example, in neuroscience, Vicente et al. \cite{vicente2011transfer} found transfer entropy to be a superior measure in detecting causality in electrophysiological communication than the auto-regressive Granger causality formulation. In climatology, Liang  derived from first principles a linear information flow measure, and used this to show that El Ni\~{n}o tends to stabilise the Indian Ocean Dipole \cite{san2014unraveling}. This analysis also detected a causal effect in the other direction; the Indian Ocean Dipole was shown to amplify El Ni\~{n}o oscillations. The technique was used with further success by Stips et al. \cite{stips2016causal} to confirm that recent CO$_2$ emissions show a one-way causality towards global mean temperature anomalies, but that on paleoclimate timescales, this direction is reversed and temperatures drive CO$_2$ levels. Finally, in finance, information transfer was measured between equities indices by Kwon \& Yang, showing that the information transfer was greatest from the US, and greatest towards the APAC region \cite{kwon2008information}. In particular, the S\&P500 was shown to be the strongest driver of other stock indices. In an earlier and somewhat related work, Marschinski \& Kantz   \cite{Marschinski2002} defined and used effective transfer entropy to quantify contagion in financial markets. Similarly, Tungsong et al. \cite{tungsong2018} used transfer entropy to develop upon the previous work by Diebold \& Yilmaz \cite{diebold2009measuring} in quantifying spillover effects between financial markets, generalising the methodology and estimating the time evolution of interconnectedness between financial systems.

 The rest of the paper is organised as follows. In Section \ref{s.background} we provide a brief background on Granger causality (linear causality measure) and transfer entropy (non-linear causality measure).  In Section \ref{s.method} we describe details of the methodology adopted to quantify and validate linear and non-linear causality, and the techniques used to generate synthetic series of linear and non-linear causal coupling. Section \ref{s.validation} demonstrates that the methodologies correctly detect causality  in the linear and non linear case when testing against synthetic data. Results for real data, concerning causality between cryptocurrency price and sentiment, are presented in Section \ref{s.results}. Section \ref{s.conclusions} reports conclusions and perspectives.

@@

  \subsection{Linear Causality}

@@

  \subsection{Non-Linear Causality}

  To detect non-linear causality, we apply an information-theoretic approach. Equation \ref{eq:geweke_measure} measures the extent to which the additional information in the lagged variable reduces the variance in the model residuals. Transfer entropy extends this concept by considering the uncertainty, instead of the variance. Adopting Shannon's measure of information \cite{shannon1948}, we can express the uncertainty associated with the continuous random variable $X$ by:

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
    Information transfer between time series is calculated using the asymmetric information-theoretic measure known as transfer entropy. Geweke's autoregressive formulation of Granger causality is used to compute linear transfer entropy, and Schreiber's general, non-parametric, information-theoretic formulation is used to quantify non-linear transfer entropy. 
    We first validate these measures against synthetic data. Then we apply these measures to detect {\color{blue} statistical causality} between social sentiment and cryptocurrency prices. We validate results by performing permutation tests by shuffling the time series, and calculate the Z-score. We also investigate different approaches for partitioning in nonparametric density estimation which can improve the significance. 
    Using these techniques on sentiment and price data over a 48-month period to August 2018, for four major cryptocurrencies, namely bitcoin (BTC), ripple (XRP), litecoin (LTC) and ethereum (ETH), we detect significant information transfer, on hourly timescales, with greater net information transfer from sentiment to price for XRP and LTC, and instead from price to sentiment for BTC and ETH. We report the scale of non-linear {\color{blue}statistical causality} to be an order of magnitude larger than {\color{blue}the linear case}.
\end{abstract}

@@

  Though developed relatively recently, information-theoretic methods have been used with success in research across disciplines, to detect information transfer where interventionist approaches are not possible. For example, in neuroscience, Vicente et al. \cite{vicente2011transfer} found transfer entropy to be a superior measure in {\color{blue} detecting Granger causality} in electrophysiological communication than {\color{blue}the auto-regressive formulation}. In climatology, Liang derived from first principles a linear information flow measure, and used this to show that El Ni\~{n}o tends to stabilise the Indian Ocean Dipole \cite{san2014unraveling}. This analysis also detected a causal effect in the other direction; the Indian Ocean Dipole was shown to amplify El Ni\~{n}o oscillations. The technique was used with further success by Stips et al. \cite{stips2016causal} to confirm that recent CO$_2$ emissions show a one-way causality towards global mean temperature anomalies, but that on paleoclimate timescales, this direction is reversed and temperatures drive CO$_2$ levels. Finally, in finance, information transfer was measured between equities indices by Kwon \& Yang, showing that the information transfer was greatest from the US, and greatest towards the APAC region \cite{kwon2008information}. In particular, the S\&P500 was shown to be the strongest driver of other stock indices. In an earlier and somewhat related work, Marschinski \& Kantz   \cite{Marschinski2002} defined and used effective transfer entropy to quantify contagion in financial markets. Similarly, Tungsong et al. \cite{tungsong2018} used transfer entropy to develop upon the previous work by Diebold \& Yilmaz \cite{diebold2009measuring} in quantifying spillover effects between financial markets, generalising the methodology and estimating the time evolution of interconnectedness between financial systems.

 The rest of the paper is organised as follows. In Section \ref{s.background} we provide a brief background on {\color{blue}Granger-Geweke causality} ({\color{blue}the} linear causality measure) and transfer entropy ({\color{blue}the} non-linear causality measure). In Section \ref{s.method} we describe details of the methodology adopted to quantify and validate linear and non-linear {\color{blue}Granger} causality, and the techniques used to generate synthetic series of linear and non-linear causal coupling. Section \ref{s.validation} demonstrates that the methodologies correctly detect {\color{blue}statitical} causality in the linear and non linear case when testing against synthetic data. Results for real data, concerning causality between cryptocurrency price and sentiment, are presented in Section \ref{s.results}. Section \ref{s.conclusions} reports conclusions and perspectives.

@@

  \subsection{Linear {\color{blue}Granger} Causality}

@@

  \subsection{Non-Linear {\color{blue}Granger} Causality}

  To detect non-linear {\color{blue}Granger} causality, we apply an information-theoretic approach. Equation \ref{eq:geweke_measure} measures the extent to which the additional information in the lagged variable reduces the variance in the model residuals. Transfer entropy extends this concept by considering the uncertainty, instead of the variance. Adopting Shannon's measure of information \cite{shannon1948}, we can express the uncertainty associated with the continuous random variable $X$ by:
